Region-network hierarchical sparsity priors for\\ high-dimensional inference in brain imaging

### 2016-01-29 paper/paper.tex
small pass on methods

% SPECIAL
Functional specialization has been
explored and interpreted based on many different research methods.
%
Single-cell recordings and microscopic examination
revealed, for instance, the
specialization in the occipital visual cortex into V1, V2, V3, V3A/B, and V4
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system,
in turn,
was frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
The whole-brain localization of
sensory, motor, and emotional functions to cortical areas
% in the living brain
was later enabled by
non-invasive brain imaging with
functional magnetic resonance imaging (fMRI) and
positron emission tomography (PET)
\cite{fristen1997imaging}.
Further,
radioactive mapping {\color{red} Michael: is this the correct term?} of neurotransmitter receptors
rendered accessible yet another
local characteristic of neuronal populations
\cite{zilles2009receptor}.
% allude to our many-regions layer in our methods
In the computational era,
automatic clustering methods are increasingly employed to
regionally differentiate the cerebral cortex,
which can partly be more fine-grained than
classical microscopical borders
\cite{behrens03, cbp2015review}.
High-thoughput approaches today enable
ultrahigh-resolution 3D models of brain anatomy
at near-cellular scale
\cite{amunts2013bigbrain}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architecture of brain organization.

@@

\section{Methods}
Our main contribution is the domain-specific adaptation of sparse structured
penalties making it possible to jointly incorporate functional 
specialization and functional integration priors into statistical 
estimators.
We make use of the hierarchical group lasso as introduced by Jennaton and 
others in \cite{jenatton2011structured} to create a set of convex penalties
with which we can probe this interplay between regional specificity and 
inter-regional functional integration during different cognitive tasks.
%
\paragraph{Rationale}
Three-dimensional voxel brain images as obtained by various neuroimaging
techniques are very high-dimensional, but also very highly structured.
While its explicit dimensionality, the number of voxels, which varies with
the resolution of the image is generally of the order of 100000, the number
of samples available for analyses never exceeds hundreds or thousands.
This \(n << p\) situation immediately implies an underdeterminedness for any
linear model basing its decisions on dot products with the voxel values.

Luckily, there is and abundance structure to be exploited, which can be
injected into the estimation as a domain knowledge prior. Here we are 
interested in incorporating knowledge about both the well-established 
modules of functional specialization and the interactions of these modules
amongst each other across the brain.
%
Developmentally, such large-scale networks emerge during late
fetal growth (Doria et al., 2010), before cognitive capacities mature
in childhood. 
%
In adults, nodes of a same cohesive network have more
similar functional profiles than nodes from different networks
(Anderson et al., 2013).
%
On the signal level, functional modules are localized structures implicating
neighboring voxels in an image. A variety of known functional modules has
been compiled into atlases of the brain. Different types of atlases rely on
different ways of segregating modules from another. For instance, anatomical
atlases rely on the variation of tissue properties across the brain, which
can be at the cellular level or based on higher-order structure.
%
Global functional networks can be discovered by statistical analysis of
functional brain imaging data, for example the BOLD contrast in fMRI. A
well-established method of estimating these networks is independent 
component analysis \cite{Hamalainen}. It robustly identifies the default
mode, saliency and attention networks.
%
Global functional networks as estimated by ICA are contiuous brain maps
attributing weights to each voxel. Since the regions involved in these
networks are nevertheless largely disjoint spatially and roughly in
correspondence with module boundaries {\color{red} Michael: verify this!!!}, 
it is possible to uniquely associate each module from a region atlas to one
of the extracted global functional networks.

This correspondence makes it possible to encode our knowledge about the 
networks and regions into a hierarchy: A brain contains a certain number of
networks, which in turn consist of the atlas modules as subregions.
We can use this hierarchy to create a structural prior of expected weight
maps for classifiers.

@@

\paragraph{Problem formulation}
We formulate our estimation problem in the framework of regularized risk
estimation applied to linear models: 
We would like to estimate a good predictor of cognitive task
given a brain image. Let the set \(\mathcal X\subset\R^p\) represent brain
images of \(p > 0\) voxels.

>>>>

% SPECIAL
Functional specialization has been
explored and interpreted based on many different research methods.
%
Single-cell recordings and microscopic examination
revealed, for instance, the
specialization in the occipital visual cortex into V1, V2, V3, V3A/B, and V4
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system,
in turn,
was frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
The whole-brain localization of
sensory, motor, and emotional functions to cortical areas
% in the living brain
was later enabled by
non-invasive brain imaging with
functional magnetic resonance imaging (fMRI) and
positron emission tomography (PET)
\cite{fristen1997imaging}.
Further,
radioactive mapping of neurotransmitter receptors
rendered accessible yet another
local characteristic of neuronal populations
\cite{zilles2009receptor}.
% allude to our many-regions layer in our methods
In the computational era,
automatic clustering methods are increasingly employed to
regionally differentiate the cerebral cortex,
which can partly be more fine-grained than
classical microscopical borders
\cite{behrens03, cbp2015review}.
High-thoughput approaches today enable
ultrahigh-resolution 3D models of brain anatomy
at near-cellular scale
\cite{amunts2013bigbrain}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architecture of brain organization.

@@

\section{Methods}
Our main contribution is the domain-specific adaptation of sparse structured
penalties making it possible to jointly incorporate functional 
specialization and functional integration priors into statistical 
estimators.
We capitalize on hierarchical group lasso as introduced by Jennaton and 
others in \cite{jenatton2011structured} to create a set of convex penalty terms.
These can probe the interplay between regional specialization and 
inter-regional functional integration during defined psychological tasks.
%
\paragraph{Rationale.}
Three-dimensional voxel brain images as obtained by various neuroimaging
techniques are very high-dimensional, but also very highly structured.
While its explicit dimensionality, the number of brain voxels, which varies with
the resolution of the image is generally of the order of 100,000, the number
of samples available for analyses rarely exceeds hundreds or thousands.
This \(n << p\) scenario immediately implies underdetermination of any
linear model based on dot products with the voxel values.

Yet, there is abundant structure to be exploited, which can be
injected into the estimation as a domain knowledge prior. Here we are 
interested in incorporating knowledge about both the well-established 
modules of functional specialization and the spatiotemporal interactions
between these.
%
Developmentally, such large-scale networks emerge during late
fetal growth (Doria et al., 2010), before mental capacities mature
in childhood. 
%
In adults, nodes of a such a spatiotemporally cohesive network have more
similar functional profiles than nodes from different networks
(Anderson et al., 2013).
%
On the signal level, functional modules are localized structures implicating
neighboring voxels in an image. A variety of known functional modules has
been compiled into atlases of the brain. Different types of atlases rely on
different ways of segregating modules from another. For instance, anatomical
atlases rely on the variation of tissue properties across the brain, which
can be at the cellular level or based on higher-order structure.
%
Global functional networks can be discovered by statistical analysis of
functional brain imaging data, for example the BOLD contrast in fMRI. A
well-established method of estimating these networks is independent 
component analysis \cite{Hamalainen}. It robustly identifies the default
mode, saliency and attention networks.
%
Global functional networks as estimated by ICA are continuous brain maps
attributing weights to each voxel. Since the regions involved in these
networks are nevertheless largely disjoint spatially and roughly in
correspondence with module boundaries {\color{red} Danilo: Ok!}, 
it is possible to uniquely associate each module from a region atlas to one
of the extracted global functional networks.

This correspondence makes it possible to inject established knowledge about the 
networks and regions into a hierarchy: A brain contains a certain number of
networks, which in turn consist of the atlas modules as subregions.
We can use this hierarchy to create a structural prior of expected weight
maps for classifiers.

@@

\paragraph{Problem formulation.}
We formulate our estimation problem in the framework of regularized risk
estimation applied to linear models: 
We would like to estimate a good predictor of cognitive task
given a brain image. Let the set \(\mathcal X\subset\R^p\) represent brain
images of \(p > 0\) voxels.
